---
title: "EDSD 2024-2025 Sequence Analysis Assignment"
author: "Gilbert Habaasa"
date: "2025-07-05"
output:
  word_document: default
  html_document: default
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)

```

#Instructor: Nicola Barban

##Assignment
Use the data provided to illustrate life course trajectories of a sample of individuals from
the German Family Panel (pairfam) data. Data come from Raab and Struffolino (2022). Additional details and code can be found in the book and in the companion github page. 

Prepare a 5 page max report in which you:
1. Import the data in R and describe the distribution of life states using relevant sequence
analysis tools, including tables, indexes and figures.
2. Calculate life course dissimilarities in the data.
3. Use a clustering methods to derive typical trajectories and describe groups

```{r}
## Load Required Packages
library(haven)        # To read Stata files
library(tidyverse)    # To enable loading of multiple packages and manipulation
library(TraMineR)     # For sequence analysis
library(RColorBrewer) # For color palettes
library(cluster)      # For clusters
```

##A. Import the data in R and describe the distribution of life states using relevant sequence
```{r}
## Step 1.1: Import Data from Stata
pairfam_raw <- read_dta("C:/Users/admin/OneDrive - London School of Hygiene and Tropical Medicine/‌INED 2024/Sequence analysis/Assignment/PartnerBirthbio.dta")
# View structure and variable names
#names(pairfam_raw)

## Step 1.2: Recode Life States
pairfam <- pairfam_raw %>%
  select(id, weight40, starts_with("state")) %>% # Keep ID, weights, and state vars
  mutate(across(starts_with("state"), ~ case_when(
    . < 3 ~ 1,            # Single
    . %in% c(3, 4) ~ 2,   # Living Apart Together (LAT)
    . %in% c(5, 6) ~ 3,   # Cohabiting
    . > 6 ~ 4             # Married
  )))

## Step 1.3: Define State Labels, Codes, and Colors
pairfam.labels <- c("Single", "Living Apart Together", "Cohabiting", "Married")
pairfam.scode  <- c("S", "LAT", "COH", "MAR")
pairfam.colors <- brewer.pal(4, "Set1")  # Use 4 colors for 4 states
pairfam$sex <- factor(pairfam_raw$sex, levels = c(0, 1), labels = c("Male", "Female"))

## Step 1.4: Create Sequence Object
# Identify sequence columns
seq_columns <- which(names(pairfam) %in% paste0("state", 1:265)) 

# Define sequence object
pairfam.seq <- seqdef(pairfam,var = seq_columns,states = pairfam.scode,labels = pairfam.labels,
                      weights = pairfam$weight40,id = pairfam$id,cpal = pairfam.colors)
# Confirm definitions
alphabet(pairfam.seq) #tell me possible alphabets
stlab(pairfam.seq) #possible colours for representation

# Step 1.5: Sequence Summary Plots: Frequency and distribution plots
seqfplot(pairfam.seq, main = "Sequence Frequency Plot",with.legend = TRUE,legend.prop = 0.3,cex.legend= 1.0)
seqdplot(pairfam.seq, main = "State Distribution Plot",with.legend = TRUE,legend.prop = 0.3,cex.legend= 1.0)

# Plot distribution plots grouped by gender
seqdplot(pairfam.seq, group = pairfam$sex,border = NA, space = 0, main = "Distribution Plot",with.legend = TRUE,legend.prop = 0.3,cex.legend= 0.8)

## Step 1.6: Statistics and Distribution of Life States

### First 10 Most Common Sequences
seqtab(pairfam.seq, idxs = 1:10) # Shows top ten most common sequences

### Mean Duration in each state
state_codes <- c("S", "LAT", "COH", "MAR")
pairfam.duration <- as.data.frame(seqistatd(pairfam.seq))
pairfam.duration$sex <- pairfam$sex

# Computed mean durations by gender
mean_durations_by_gender <- pairfam.duration %>%
  group_by(sex) %>%
  summarise(across(all_of(state_codes), ~ mean(.x, na.rm = TRUE)))
print(mean_durations_by_gender)

# Mean time in each state by Gender
seqmtplot(pairfam.seq, group = pairfam$sex, main = "Mean time", with.legend = TRUE,legend.prop = 0.3,cex.legend= 0.8)

### Transition Matrix
pairfam.trate <- seqtrate(pairfam.seq)
round(pairfam.trate, 2)

### Entropy Over Time
entropy <- seqstatd(pairfam.seq)$Entropy
plot(entropy,type = "l",col = "black",main = "Entropy of the State Distribution",
     xlab = "Time (months)", ylab = "Entropy")

### Sequence Turbulence
turbulence <- seqST(pairfam.seq)
hist(turbulence,col = "grey", main = "Histogram of Sequence Turbulence",
     xlab = "State",ylab = "Sequence Turbulence")
```
**The most common sequence is Singlehood state and is followed by sequences that start with singlehood and transition to LAT, cohabitation and marriage On average, males spend 80.1 months in single state while females spend 48.3 months in single state. However, in marriage state, the males spend less time than females (88.4 months vs 120.0 months).On transitions,persons single have 98% chance of staying single. The 98% people stay in LAT state while 100% remain married.On theother hand, the 98% cohabiting are likely to keep in  same state.**

##2. Calculate life course dissimilarities in the data.

```{r}

#Step 2.1: Compute dissimilarities using Optimal Matching (OM) with constant substitution costs

pairfam.dist1 <- seqdist(pairfam.seq, method = "OM", indel = 1, sm = "CONSTANT")
#pairfam.dist1[1:10, 1:10]

#Step 2.2: Compute dissimilarities using Optimal Matching (OM) with transition rate substitution matrix
submat <- seqsubm(pairfam.seq, method = "TRATE")
pairfam.dist2 <- seqdist(pairfam.seq, method = "OM", indel = 1, sm = submat)
#pairfam.dist2[1:10, 1:10]

#Step 2.3: Compute dissimilarities using Optimal Matching (OM) with empirically derived insertion/deletion and substitution costs
trcost <- seqcost(pairfam.seq, method="TRATE")
pairfam.dist3=seqdist(pairfam.seq,method="OM",indel=trcost$indel,sm=trcost$sm)
#pairfam.dist3[1:10,1:10]

#Step 2.4: Compute dissimilarities using Longest Common Subsequence (LCS) and Dynamic Hamming Distance (DHD) Matrices
dist.lcs=seqdist(pairfam.seq, method="LCS")
dist.dhd=seqdist(pairfam.seq, method="DHD")

#Step 2.5: Computed dissimilarity matrix
dissimilarities <- list(OM_constant = pairfam.dist1,OM_trate = pairfam.dist2,
  OM_seqcost = pairfam.dist3,LCS = dist.lcs,DHD = dist.dhd)

#Step 2.6: Comparison of Dissimilarity matrices using Pairwise Correlation

dissim_df <- data.frame(OM_constant = c(pairfam.dist1),OM_trate = c(pairfam.dist2),
  OM_seqcost  = c(pairfam.dist3),LCS= c(dist.lcs),DHD= c(dist.dhd))
cor(dissim_df)

```
**There is high correlation between OM,LCS and DHD. This implies similar results despite the method used.**

##3. Use a clustering methods to derive typical trajectories and describe groups

```{r}
  # Step 3.1: Clustering using Ward's method
pairfam.clusterward <- agnes(pairfam.dist3, diss = TRUE, method = "ward")
plot(pairfam.clusterward, which.plots = 2, main = "Dendrogram of agnes", labels = FALSE)

# Step 3.2: Chosing of 3 clusters
pairfam.cl3 <- cutree(pairfam.clusterward, k = 3)

# Step 3.3: Convert to factor and assign descriptive labels
pairfam.cl3 <- factor(pairfam.cl3,levels = 1:3,
                      labels = c("Mostly Single/Late Partn",
                                 "Early Cohab/then Marr", 
                                 "Mixed LAT/Less Marr")) #"Mostly Single/Late Partnership",
                                 #"Early Cohabitation/then Marriage", "Mixed Living Apart Together/Less Marriage".

# Step 3.4: Add cluster labels to your dataset
pairfam$cluster_label <- pairfam.cl3

#Step 3.5: Generate different cluster plots
par(mar = c(5, 4, 4, 4))
seqdplot(pairfam.seq, group = pairfam.cl3,
         border = NA,space = 0)

# Mean time spent in each state
par(mar = c(5, 4, 4, 4))
seqmtplot(pairfam.seq,group = pairfam.cl3,
          border = NA, space = 0)
```

**Based on 3 clusters, a diverse life course trajectory in partnership formation is observed: stable singlehood or delayed partnership; early cohabitation leading to marriage; mixed living apart together and less marriage. The first cluster is dominated by singlehood sequence. The second cluster is largely dominated by early cohabitation and marriage in the longrun. Both early cohabiting and mixed LAT report partners staying long in marriage in the long run.**  
